---
title: "RNA-seq analysis"
author: "Mark Lopez"
date: "2026-04-24"
output: html_document
---

```{r setup, include=FALSE}
install.packages("BiocManager")
install.packages("gplots")
BiocManager::install("RNAseq123")
library(limma)
library(Glimma)
library(edgeR)
library(Mus.musculus)

```
This section Installs and loads bioconducter packages that are being used in this project/workflow for RNA-seq analysis. The limma package is used for linear modeling, differential expression, and plotting data. The Glimma packag is used to generate interactive plots to visualize gene expression and differential expression results. The edgeR package is used to read RNA-seq data utilizing statistical models for differential expressed genes, filtering low expressed genes and normalizing the data. The Mus.musculus loads the mouse annotation package for mapping and labeling data, gene ID to gene symbols and chromosome information.

```{r}
#Data packaging reading in count data
url <- "https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE63310&format=file"
utils::download.file(url, destfile="GSE63310_RAW.tar", mode="wb") 
utils::untar("GSE63310_RAW.tar", exdir = ".")
files <- c("GSM1545535_10_6_5_11.txt", "GSM1545536_9_6_5_11.txt", "GSM1545538_purep53.txt",
  "GSM1545539_JMS8-2.txt", "GSM1545540_JMS8-3.txt", "GSM1545541_JMS8-4.txt",
  "GSM1545542_JMS8-5.txt", "GSM1545544_JMS9-P7c.txt", "GSM1545545_JMS9-P8c.txt")
for(i in paste(files, ".gz", sep=""))
  R.utils::gunzip(i, overwrite=TRUE)

files <- c("GSM1545535_10_6_5_11.txt", "GSM1545536_9_6_5_11.txt", 
   "GSM1545538_purep53.txt", "GSM1545539_JMS8-2.txt", 
   "GSM1545540_JMS8-3.txt", "GSM1545541_JMS8-4.txt", 
   "GSM1545542_JMS8-5.txt", "GSM1545544_JMS9-P7c.txt", 
   "GSM1545545_JMS9-P8c.txt")
read.delim(files[1], nrow=5)

x <- readDGE(files, columns=c(1,3))
class(x)
```

In this portion we are first retrieving GEO download url link containing the data and storing it in the url variable. Then we are downloading the file from GEO link and naming it GSE63310_RAW.tar and making it in binary mode. Saving the file in binary mode is ensuring the tar file is saved without any conversion errors or corruption caused. Then the files gets unpacked so the count files become available in the working directory. The code then stores the txt.files into a variable called files utilizing the c() so the txt.files get combined into a vector and then in the following line of code it creates a for loop going through each files, decompressing (gunzipping them) the files and storing the files as regular txt.files so R can read them properly. This portion also has a line of code that lets you check the preview of the structure, with eh ID, gene length and count before importing all the files. Then the last part of code in this section imports all gene IDs and count the values from all 9 samples into a single object for analysis and is confirmed that it was loaded into a edgeR DGEList.

```{r}
#organizing sample information 
samplenames <- substring(colnames(x), 12, nchar(colnames(x)))
samplenames
colnames(x) <- samplenames
group <- as.factor(c("LP", "ML", "Basal", "Basal", "ML", "LP", 
                     "Basal", "ML", "LP"))
x$samples$group <- group
lane <- as.factor(rep(c("L004","L006","L008"), c(3,4,2)))
x$samples$lane <- lane
x$samples
```
In this section of the code it cleans up the sample names so there are easier to read, so instead of using the GEO names they get shorter names so it's easier for the plots and tables. This section also sets the biological condition for each sample and also attaches the biological group to the RNA-seq object so it knows which sample data belongs to which biological group ("LP","ML","Basal").This is essential because we need to compare the groups to get the differentially expression analysis. This section also stores sample to a sequencing lane so it could have a batch variable and could be used for a technical variation in statistical models. Now each samples has both biological information and technical batch information to support the model to separate the biological differences from lane related effects.

```{r}
#organizing gene annotations 
geneid <- rownames(x)
genes <- select(Mus.musculus, keys=geneid, columns=c("SYMBOL", "TXCHROM"), 
                keytype="ENTREZID")
head(genes)
genes <- genes[!duplicated(genes$ENTREZID),]
x$genes <- genes
x
```
This chunks of code organizes gene annotations by extracting the gene IDs from the count matrix and are used to look up gene and chromosome information. It also turns the numerical gene IDs into more useful information. It only adds one annotation per gene ID as different gene IDs could have more than one annotation. Then the gene annotation table is added to the DGEList object so that the RNA-seq object contains raw counts, sample metadata, and gene annotations and ends with outputting the full DGEList containing samples, counts, and gene annotation. 

```{r}
#Data preprocesssing
#transforming from raw scale
cpm <- cpm(x)
lcpm <- cpm(x, log=TRUE)
L <- mean(x$samples$lib.size) * 1e-6
M <- median(x$samples$lib.size) * 1e-6
c(L, M)
summary(lcpm)
```
This section converts raw counts into counts per million "CPM" in order to put the samples on a more comparable scale. It then takes the CPM and log2 transform's the CPM so the data is easier to visualize across samples. It also produces the average library size and median library size in millions to better understand what the ranges for the cutoff in the dataset, then printing both average and median library sizes together along with summarizing the log counts per million values for each samples. This shows an overview of the expression distribution in each sample. 

```{r}
#removing genes that are lowly expressed
table(rowSums(x$counts==0)==9)
keep.exprs <- filterByExpr(x, group=group)
x <- x[keep.exprs,, keep.lib.sizes=FALSE]
dim(x)
lcpm.cutoff <- log2(10/M + 2/L)
library(RColorBrewer)
nsamples <- ncol(x)
col <- brewer.pal(nsamples, "Paired")
par(mfrow=c(1,2))
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="A. Raw data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")
lcpm <- cpm(x, log=TRUE)
plot(density(lcpm[,1]), col=col[1], lwd=2, ylim=c(0,0.26), las=2, main="", xlab="")
title(main="B. Filtered data", xlab="Log-cpm")
abline(v=lcpm.cutoff, lty=3)
for (i in 2:nsamples){
den <- density(lcpm[,i])
lines(den$x, den$y, col=col[i], lwd=2)
}
legend("topright", samplenames, text.col=col, bty="n")

```
This next section of the code removes genes that are lowly expressed. It starts off by creating a count of the amount of genes that have 0 counts in all 9 samples that are not useful in the differentially expression analysis. It then decides which genes are expressed enough to keep through filtering by expressing to avoid adding noise and keeping the express genes that meet the threshold of expression. As a check point, It prints the number of filtered genes and the sample size. T

```{r}
#Normalizing gene expression distributions
x <- calcNormFactors(x, method = "TMM")
x$samples$norm.factors
x2 <- x
x2$samples$norm.factors <- 1
x2$counts[,1] <- ceiling(x2$counts[,1]*0.05)
x2$counts[,2] <- x2$counts[,2]*5
par(mfrow=c(1,2))
lcpm <- cpm(x2, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="A. Example: Unnormalised data",ylab="Log-cpm")
x2 <- calcNormFactors(x2)  
x2$samples$norm.factors
lcpm <- cpm(x2, log=TRUE)
boxplot(lcpm, las=2, col=col, main="")
title(main="B. Example: Normalised data",ylab="Log-cpm")
```
This part of the project normalizes gene expression distributions by filtering out data counts using TMM normalization.The code normalizes the dataset so samples are more comparable across the different gene expression distributions. It helps correct cases where one sample may have more highly expressed genes than another sample. The workflow shows how much each sample had to adjsuted during normalization and it creates a copy so it can demonstrate the normalization without affecting the actual data set. Overall, this section makes the samples more comparable before clustering and differential expression testing.

```{r}
#Unsupervised clustering of samples
lcpm <- cpm(x, log=TRUE)
par(mfrow=c(1,2))
col.group <- group
levels(col.group) <-  brewer.pal(nlevels(col.group), "Set1")
col.group <- as.character(col.group)
col.lane <- lane
levels(col.lane) <-  brewer.pal(nlevels(col.lane), "Set2")
col.lane <- as.character(col.lane)
plotMDS(lcpm, labels=group, col=col.group)
title(main="A. Sample groups")
plotMDS(lcpm, labels=lane, col=col.lane, dim=c(3,4))
title(main="B. Sequencing lanes")
glMDSPlot(lcpm, labels=paste(group, lane, sep="_"), 
          groups=x$samples[,c(2,5)], launch=FALSE)
```
This section of the code checks if the samples are clustering in a reasonable way before the actual differential expression analysis. The first MDS plots to show if the samples are from the the same biological group since each sample group should be clustering with the same biological group such as basal with basal. The second MDS plot checks if the the sequencing lane has any influence on the data that could suggest a potential batch effect. Overall this section is a good checkpoint to visually check if the experiment has a strong biological signal and whether technical effects need to be included in the model

```{r}
#Differential expression analysis
#creating a design matix and contrasts

design <- model.matrix(~0+group+lane)
colnames(design) <- gsub("group", "", colnames(design))
design


contr.matrix <- makeContrasts(
   BasalvsLP = Basal-LP, 
   BasalvsML = Basal - ML, 
   LPvsML = LP - ML, 
   levels = colnames(design))
contr.matrix

```
This section goes into the differential expression analysis by creating a design matrix and contrast matrix. The design matrix tells limma what variables to include in the model like the biological and the sequencing lane. This allows the analysis to compare cells types while also including any possible lane effects. The contrast matrix tells limma which specific group comparisons to test. The comparison groups being Basal vs LP, Basal vs ML and LP vs ML.

```{r}
#Removing heteroscedascity from count data
par(mfrow=c(1,2))
v <- voom(x, design, plot=TRUE)
v

vfit <- lmFit(v, design)
vfit <- contrasts.fit(vfit, contrasts=contr.matrix)
efit <- eBayes(vfit)
plotSA(efit, main="Final model: Mean-variance trend")

```

This next part uses voom() to prepare RNA-seq count data for limma analysis. The RNA-seq data has heteroscedasticity, meaning the variance of low expression and high expression behave differently. The voom() estimates this mean-variance pattern and creates weights to help the model account for the variances and this makes the data more suitable to for linear modeling with limma. This bottom portion of this section performs the main statistical modeling. First, lmFit() fits a model to each gene using the voom-transformed data. Then contrasts.fit() applies the specific comparisons that were defined in the contrast matrix. Next, eBayes() improves the reliability of the statistical results by using the information across genes. The final plot checks whether the model has handled the mean-variance trend sufficiently

```{r}
#Examining the number of DE genes
summary(decideTests(efit))
tfit <- treat(vfit, lfc=1)
dt <- decideTests(tfit)
summary(dt)

de.common <- which(dt[,1]!=0 & dt[,2]!=0)
length(de.common)

head(tfit$genes$SYMBOL[de.common], n=20)

vennDiagram(dt[,1:2], circle.col=c("turquoise", "salmon"))

write.fit(tfit, dt, file="results.txt")

```

This section summarizes how many genes are differential expressed in each group comparison. It first uses the Bayes results, then applies a fold change test with treat(). The code also identifies genes that are significant in more than one comparison, which can reveal shared biological patterns. The Venn diagram gives a visual summary of overlap between comparisons. Finally, the results are saved to results.txt, which becomes one of the main output files from the analysis.


```{r}
#Examining individual DE genes from top to bottom
basal.vs.lp <- topTreat(tfit, coef=1, n=Inf)
basal.vs.ml <- topTreat(tfit, coef=2, n=Inf)
head(basal.vs.lp)

head(basal.vs.ml)


```
This section of the code inspects individual differential expressed genes. topTreat() creates ranked tables so you can see which genes are the strongest hits in each comparison. These tables include both statistical information and gene annotation information. This is where we can identifying specific genes that may be biologically important. It is useful for deciding which genes to discuss in a report or presentation or to research more into. 
```{r, fig.width=10, fig.height=10}
#Useful graphical representations of differential expression results
plotMD(tfit, column=1, status=dt[,1], main=colnames(tfit)[1], 
       xlim=c(-8,13))

glMDPlot(tfit, coef=1, status=dt, main=colnames(tfit)[1],
         side.main="ENTREZID", counts=lcpm, groups=group, launch=FALSE)


library(gplots)
basal.vs.lp.topgenes <- basal.vs.lp$ENTREZID[1:100]
i <- which(v$genes$ENTREZID %in% basal.vs.lp.topgenes)
mycol <- colorpanel(1000,"blue","white","red")
heatmap.2(lcpm[i,], scale="row",
   labRow=v$genes$SYMBOL[i], labCol=group, 
   col=mycol, trace="none", density.info="none", 
   margin=c(8,6), lhei=c(2,10), dendrogram="column")


```
This section provides a plot to visualizes the differential expression results. The  MD plot shows how log fold change relates to average expression across the different genes. The Glimma plot makes this interactive, allowing individual genes to be explored more easily. The heatmap focuses on the top 100 Basal vs LP genes and shows their expression patterns across all samples. Together, these plots help confirm whether the statistical results make biological sense and whether samples cluster as expected.

```{r}
#Gene set testing with camera
library(RNAseq123)
load(system.file("extdata", "mouse_c2_v5p1.rda", package = "RNAseq123"))
idx <- ids2indices(Mm.c2,id=rownames(v))
cam.BasalvsLP <- camera(v,idx,design,contrast=contr.matrix[,1])
head(cam.BasalvsLP,5)
cam.BasalvsML <- camera(v,idx,design,contrast=contr.matrix[,2])
head(cam.BasalvsML,5)
cam.LPvsML <- camera(v,idx,design,contrast=contr.matrix[,3])
head(cam.LPvsML,5)
barcodeplot(efit$t[,3], index=idx$LIM_MAMMARY_LUMINAL_MATURE_UP, 
            index2=idx$LIM_MAMMARY_LUMINAL_MATURE_DN, main="LPvsML")

```

This section shifts towards the analysis from individual genes to groups of related genes. Instead of looking at only which single genes are differential expressed, camera() focuses on whether entire biological pathways or gene signatures are enriched. This can provide more biological interpretation because pathways are easier to understand than long gene lists. The workflow runs gene set testing for all three comparisons: Basal vs LP, Basal vs ML, and LP vs ML. The barcode plot then gives a visual example of where specific mature luminal gene sets apear in the ranked results. 

```{r}
#Software and code use
sessionInfo()
```
 
This section records the software environment used to run the analysis. sessionInfo() shows the R version and package versions. This is an important because different versions can give different results. This helps if someone else wants to rerun the workflow, this information helps them match the original setup. It also helps troubleshoot issues if code works on one computer but not another. In a report or lab submission, this section supports transparency about how the analysis was performed.
